---
title: "Feature Selection: Practical 3"
subtitle: "Overfitting to Features"
author:
 - name: Solon Karapanagiotis
   affiliations: MRC Biostatistics Unit, University of Cambridge
date: "21-04-2026"
output:
    html_document:
        theme: cosmo
        highlight: tango
        toc: true
        toc_float: true
        number_sections: true
        df_print: paged
---

# Introduction
Explore the impact of overfitting to feature selection. 

# Dataset
The same as in Practical 2. It contains:

- 500 observations,
- 100 features (X1–X100), and 
- a binary outcome $y \in \{0,1\}$.

```{r}
path_to_data <- "/Users/solon/Cloud-Drive/Teaching/MPhil_Data_Science/Applied_ML/feature_selection/data/"

load(paste0(path_to_data, "data_practical2.Rdata"))
```

```{r}
str(data)
```



```{r, message=F, warning=F}
library(randomForest)
library(pROC)

set.seed(42)

# ============================================================
# PARAMETERS
# ============================================================
subset_sizes <- 100:1 # feature subset sizes defined here

# ============================================================
# 1. Split data into training (analysis) and test set
# ============================================================
n <- nrow(data)
train_idx <- sample(1:n, size = 0.7 * n)

train_data <- data[train_idx, ]
test_data  <- data[-train_idx, ]
```

# Approach 1 

1.  Rank predictors using the training set.
2.  For feature subset sizes from **100 down to 1**:
    1.  For each resample:
        -   Fit the model using the feature subset on the **analysis set**.
        -   Predict the **assessment set**.
    2.  End resampling loop.
3. Determine the best feature subset using resampled performance.
4. Fit the best feature subset using the **entire training set**.

# Approach 2

1.  Split data into **analysis** and **assessment** sets.
2.  For each resample:
    1.  Rank predictors using the **analysis set**.
    2.  For feature subset sizes from **100 down to 1**:
        -   Fit the model using the feature subset on the **analysis set**.
        -   Predict the **assessment set**.
    3.  End feature subset loop.
3.  Average the resampled performance for each model and feature subset size.
4.  Fit the best feature subset using the **entire training set**.
5.  Choose the model and feature subset with the best performance.
6.  Fit the final model using the best feature subset on the **entire training set**.

