---
title: "Feature Selection: Practical 2"
subtitle: "Comparing Feature Selection Methods"
author:
 - name: Solon Karapanagiotis
   affiliations: MRC Biostatistics Unit, University of Cambridge
date: "21-04-2026"
output:
    html_document:
        theme: cosmo
        highlight: tango
        toc: true
        toc_float: true
        number_sections: true
        df_print: paged
---

© `r format(Sys.Date(), "%Y")` Solon Karapanagiotis. All rights reserved.

High-dimensional binary classification problem to test different feature selection methods.

# Setup

```{r, warning=F, message=F}
library(caret)
library(pROC)
```

# Dataset
The dataset contains:

- 500 observations,
- 100 features (X1–X100), and 
- a binary outcome $y \in \{0,1\}$.

```{r}
load(here::here("data", "data_practical2.Rdata"))
```

```{r}
str(data)
```

This code splits the dataset into training and test sets using a stratified sampling approach.

First, `set.seed(1)` ensures the split is reproducible. The function `createDataPartition()` from the caret package creates a partition of the outcome variable $y$, selecting 70% of the observations for training while preserving the original class distribution. This is important in binary classification to avoid imbalance between classes in the split.

```{r}
set.seed(1)
trainIndex <- caret::createDataPartition(data$y, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]
```

# Feature Selection

The `glm()` fits a logistic regression using all predictors in the training set to predict $y$. 

- `glm(...)` stands for Generalized Linear Model in R.
- `y ~ .` means the model predicts the outcome variable $y$ using all other variables in the dataset as features.
- `data = train` specifies that the model is trained on the training set.
- `family = "binomial"` indicates that this is a logistic regression, appropriate for a binary outcome (0/1).

This serves as a simple benchmark model with no feature selection.

```{r}
baseline_model <- glm(y ~ ., data = train, family = "binomial")

# generate predicted probabilities for the test set
pred <- predict(baseline_model, test, type = "response")

# model performance is evaluated using the Area Under the ROC Curve (AUC)
baseline_auc <- pROC::roc(test$y, pred, direction = "<")$auc
baseline_auc
```
## Filter Methods 

- Implement two filter methods and compare their results.

## Wrapper methods

- Implement two wrapper methods and compare their results.

## Embedded methods

- Implement two embedded methods and compare their results.

# Discussion

- Which method achieved best predictive performance?

- Which features appear in every method?


