---
title: "Feature Selection: Practical 3"
subtitle: "Overfitting to Features"
author: "Solon Karapanagiotis, MRC Biostatistics Unit, University of Cambridge"
date: "21-04-2026"
output:
    html_document:
        theme: cosmo
        highlight: tango
        toc: true
        toc_float: true
        number_sections: true
        df_print: paged
---

# Introduction
explore the impact of overfitting to feature selection

# Dataset
The dataset contains:

- 500 observations,
- 100 features (X1–X100), and 
- a binary outcome $y \in \{0,1\}$.

```{r}
path_to_data <- "/Users/solon/Cloud-Drive/Teaching/MPhil_Data_Science/Applied_ML/code/demos/out/"

load(paste0(path_to_data, "data_practical2.Rdata"))
```

```{r}
str(data)
```



```{r, message=F, warning=F}
library(randomForest)
library(pROC)

set.seed(42)

# ============================================================
# PARAMETERS
# ============================================================
subset_sizes <- 100:1 # feature subset sizes defined here

# ============================================================
# 1. Split data into training (analysis) and test set
# ============================================================
n <- nrow(data)
train_idx <- sample(1:n, size = 0.7 * n)

train_data <- data[train_idx, ]
test_data  <- data[-train_idx, ]
```

# Approach 1 

1.  Rank predictors using the training set.
2.  For feature subset sizes from **5 down to 1**:
    1.  For each resample:
        -   Fit the model using the feature subset on the **analysis set**.
        -   Predict the **assessment set**.
    2.  End resampling loop.
3. Determine the best feature subset using resampled performance.
4. Fit the best feature subset using the **entire training set**.

```{r}
# ============================================================
# 2. Rank predictors on full training set
# ============================================================
x_train_all <- train_data[, setdiff(names(train_data), "y")]
y_train_all <- train_data$y

rf_rank <- randomForest(
  x = x_train_all,
  y = as.factor(y_train_all),
  importance = TRUE
)

imp <- importance(rf_rank, type = 2)[, 1]
ranked_features <- names(sort(imp, decreasing = TRUE))

# keep max needed features only
max_k <- max(subset_sizes)
top_features <- ranked_features[1:max_k]

# ============================================================
# 3. CV setup
# ============================================================
K <- 5
fold_id <- sample(rep(1:K, length.out = nrow(train_data)))

get_auc <- function(obs, prob) {
  as.numeric(roc(obs, prob, quiet = TRUE)$auc)
}

# store results dynamically
results <- setNames(rep(NA, length(subset_sizes)), subset_sizes)

# ============================================================
# 4. Loop over subset sizes (PARAMETERIZED)
# ============================================================
for (k in subset_sizes) {

  fold_auc <- c()

  # ----------------------------------------------------------
  # 5. Cross-validation loop
  # ----------------------------------------------------------
  for (f in 1:K) {

    analysis_set   <- train_data[fold_id != f, ]
    assessment_set <- train_data[fold_id == f, ]

    x_a <- analysis_set[, setdiff(names(train_data), "y"), drop = FALSE]
    y_a <- analysis_set$y

    x_b <- assessment_set[, setdiff(names(train_data), "y"), drop = FALSE]
    y_b <- assessment_set$y

    # ensure valid subset
    features_k <- top_features[1:min(k, length(top_features))]

    if (length(features_k) < k) next

    # ------------------------------------------------------
    # 6. Fit model
    # ------------------------------------------------------
    model <- randomForest(
      x = x_a[, features_k, drop = FALSE],
      y = as.factor(y_a),
      ntree = 500
    )

    # ------------------------------------------------------
    # 7. Predict
    # ------------------------------------------------------
    prob <- predict(
      model,
      x_b[, features_k, drop = FALSE],
      type = "prob"
    )[, 2]

    fold_auc <- c(fold_auc, get_auc(y_b, prob))
  }

  # ----------------------------------------------------------
  # 8. Store mean performance
  # ----------------------------------------------------------
  results[as.character(k)] <- mean(fold_auc, na.rm = TRUE)
}

# ============================================================
# 9. Select best subset size
# ============================================================
best_k <- as.numeric(names(which.max(results)))
best_features <- top_features[1:best_k]

cat("Best subset size:", best_k, "\n")

# ============================================================
# 10. Fit final model on full training set
# ============================================================
final_model <- randomForest(
  x = x_train_all[, best_features, drop = FALSE],
  y = as.factor(y_train_all),
  ntree = 500
)

# ============================================================
# 11. Evaluate on test set
# ============================================================
test_prob <- predict(
  final_model,
  test_data[, best_features, drop = FALSE],
  type = "prob"
)[, 2]

final_auc <- get_auc(test_data$y, test_prob)

cat("Final test AUC:", final_auc, "\n")
```

## Approach 2

1.  Split data into **analysis** and **assessment** sets.
2.  For each resample:
    1.  Rank predictors using the **analysis set**.
    2.  For feature subset sizes from **5 down to 1**:
        -   Fit the model using the feature subset on the **analysis set**.
        -   Predict the **assessment set**.
    3.  End feature subset loop.
3.  Average the resampled performance for each model and feature subset size.
4.  Fit the best feature subset using the **entire training set**.
5.  Choose the model and feature subset with the best performance.
6.  Fit the final model using the best feature subset on the **entire training set**.

```{r}
# ============================================================
# Helper function
# ============================================================
get_auc <- function(obs, prob) {
  as.numeric(roc(obs, prob, quiet = TRUE)$auc)
}

# ============================================================
# 2. Create resampling folds (analysis / assessment)
# ============================================================
K <- 5
fold_id <- sample(rep(1:K, length.out = nrow(train_data)))

# ============================================================
# 3. Storage for results
# ============================================================
results <- numeric(length(subset_sizes))
names(results) <- as.character(subset_sizes)

# ============================================================
# 4. Loop over feature subset sizes
# ============================================================
for (k in subset_sizes) {

  fold_auc <- c()

  # ==========================================================
  # For each resample (CV loop)
  # ==========================================================
  for (f in 1:K) {

    # --------------------------------------------------------
    # Split into analysis / assessment
    # --------------------------------------------------------
    analysis_set   <- train_data[fold_id != f, ]
    assessment_set <- train_data[fold_id == f, ]

    x_a <- analysis_set[, setdiff(names(train_data), "y"), drop = FALSE]
    y_a <- analysis_set$y

    x_b <- assessment_set[, setdiff(names(train_data), "y"), drop = FALSE]
    y_b <- assessment_set$y

    # --------------------------------------------------------
    # Rank predictors using ANALYSIS set only
    # --------------------------------------------------------
    rf_rank <- randomForest(
      x = x_a,
      y = as.factor(y_a),
      importance = TRUE
    )

    imp <- importance(rf_rank, type = 2)[, 1]
    ranked <- names(sort(imp, decreasing = TRUE))

    features_k <- ranked[1:min(k, length(ranked))]

    # --------------------------------------------------------
    # Fit model on analysis set
    # --------------------------------------------------------
    model <- randomForest(
      x = x_a[, features_k, drop = FALSE],
      y = as.factor(y_a),
      ntree = 500
    )

    # --------------------------------------------------------
    # Predict assessment set
    # --------------------------------------------------------
    prob <- predict(
      model,
      x_b[, features_k, drop = FALSE],
      type = "prob"
    )[, 2]

    fold_auc <- c(fold_auc, get_auc(y_b, prob))
  }

  # ----------------------------------------------------------
  # Average performance for subset size k
  # ----------------------------------------------------------
  results[as.character(k)] <- mean(fold_auc)
}

# ============================================================
# 5. Select best subset size
# ============================================================
best_k <- as.numeric(names(which.max(results)))
cat("Best subset size:", best_k, "\n")

# ============================================================
# 6. Final ranking on full training set
# ============================================================
rf_final_rank <- randomForest(
  x = train_data[, setdiff(names(train_data), "y")],
  y = as.factor(train_data$y),
  importance = TRUE
)

imp_final <- importance(rf_final_rank, type = 2)[, 1]
ranked_final <- names(sort(imp_final, decreasing = TRUE))

best_features <- ranked_final[1:best_k]

# ============================================================
# 7. Fit final model on full training set
# ============================================================
final_model <- randomForest(
  x = train_data[, best_features, drop = FALSE],
  y = as.factor(train_data$y),
  ntree = 500
)

# ============================================================
# 8. Evaluate on test set
# ============================================================
test_prob <- predict(
  final_model,
  test_data[, best_features, drop = FALSE],
  type = "prob"
)[, 2]

final_auc <- get_auc(test_data$y, test_prob)

cat("Final test AUC:", final_auc, "\n")
```


