---
title: "Feature Selection: Practical 2"
subtitle: "Comparing Feature Selection Methods"
author:
 - name: Solon Karapanagiotis
   affiliations: MRC Biostatistics Unit, University of Cambridge
date: "21-04-2026"
output:
    html_document:
        theme: cosmo
        highlight: tango
        toc: true
        toc_float: true
        number_sections: true
        df_print: paged
---

© `r format(Sys.Date(), "%Y")` Solon Karapanagiotis. All rights reserved.

High-dimensional binary classification problem designed to test different feature selection methods.

# Setup

```{r, message=F}
library(caret)
library(pROC)
library(dplyr)
```

# Dataset

The dataset contains:

- 500 observations,
- 100 features (X1–X100),  
- a binary outcome $y \in \{0,1\}$.

```{r}
load(here::here("data", "data_practical2.Rdata"))
```

```{r}
str(data)
```

This code splits the dataset into training and test sets using a stratified sampling approach.

First, `set.seed(1)` ensures the split is reproducible. The function `createDataPartition()` from the caret package creates a partition of the outcome variable y, selecting 70% of the observations for training while preserving the original class distribution. This is important in binary classification to avoid imbalance between classes in the split.

```{r}
set.seed(1)
trainIndex <- caret::createDataPartition(data$y, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]
```

# Feature Selection

The `glm()` fits a logistic regression using all predictors in the training set to predict $y$. This serves as a simple benchmark model with no feature selection.

```{r}
baseline_model <- glm(y ~ ., data = train, family = "binomial")

# generate predicted probabilities for the test set
pred <- predict(baseline_model, test, type = "response")

# model performance is evaluated using the Area Under the ROC Curve (AUC)
baseline_auc <- pROC::roc(test$y, pred, direction = "<")$auc
baseline_auc
```

## Filter Methods 

### Univariate tests based on p-value

```{r}
# Univariate tests based on p-value
pvals <- sapply(train[,-1], function(x){
  summary(glm(train$y ~ x, family = "binomial"))$coeff[2,4]
})

filter_features <- names(sort(pvals))[1:10]
filter_features
```

```{r}
formula_filter <- as.formula(
    paste("y ~", paste(filter_features, collapse = "+"))
)

filter_model <- glm(formula_filter, data = train, family = "binomial")

pred <- predict(filter_model, test, type = "response")
filter_auc <- pROC::roc(test$y, pred)$auc
filter_auc
```

> Note: For our purposes, the p-values are being used algorithmically to search for appropriate subsets, especially since it is a convenient quantity to use when dealing with different feature types. It is very important to not take p-values generated in such processes as meaningful or formal indicators of statistical significance.

### Mutual Information

Mutual information (MI) is a concept from information theory that measures the dependence between two variables. It quantifies how much knowing one variable reduces uncertainty about another.

#### Intuition

Mutual information answers the question:

> How much information about Y do we gain by observing X?

- If X and Y are independent MI = 0  
- If X strongly predicts Y MI is high  
- If X partially relates to Y MI is between 0 and high

#### Mathematical definition

MI between X and Y is defined as:

$$ I(X;Y) = \sum_{x \in X} \sum_{y \in Y} p(x,y)\log\left(\frac{p(x,y)}{p(x)p(y)}\right)$$

This compares:

- the joint distribution $p(x,y)$
- with the distribution under independence $p(x)p(y)$

An equivalent form is

$$ I(X;Y) = H(Y) - H(Y|X)$$ 

where:

- $H(Y)$ is the entropy (uncertainty) of Y
- $H(Y|X)$ is the remaining uncertainty in Y after observing X

MI measures the reduction in uncertainty about the outcome after observing a feature.

- Higher MI → more informative feature  
- Lower MI → less informative feature  

```{r}
# Mutual Information
library(FSelectorRcpp)

info_gain <- information_gain(y ~ ., data = train)

head(info_gain)

# Sort variables by importance:
mi_features <- info_gain %>%
  arrange(desc(importance)) %>%
  slice(1:10) %>%
  pull(attributes)

mi_features

# Train model using MI-selected features 
formula_mi <- as.formula(
  paste("y ~", paste(mi_features, collapse = "+"))
)

mi_model <- glm(formula_mi, data = train, family = "binomial")

pred <- predict(mi_model, test, type = "response")
mi_auc <- pROC::roc(test$y, pred)$auc
mi_auc
```

## Wrapper methods

### Stepwise regression

```{r}
# Stepwise regression
step_model <- step(glm(y ~ ., data = train, family = "binomial"),
                   direction = "both", trace = 0)

summary(step_model)$coefficients
```
```{r}
pred <- predict(step_model, test, type = "response")
step_auc <- pROC::roc(test$y, pred)$auc
step_auc
```

### Recursive Feature Elimination (RFE)

First, we define the RFE control setting using `rfeControl()`: RFE will repeatedly train logistic regression models and evaluate them using 5-fold CV. We also need to specify how many features RFE should try: models with 5, 10, 15, 20, and 30 features. The algorithm will determine which size gives the best performance. We then run the RFE (`rfe()`).

Algorithm: For each subset size (e.g., 5, 10, ...):

1. Fit a logistic regression model
2. Rank variables by importance
3. Remove the least important features
4. Evaluate performance using cross-validation
5. Store the best-performing subset

```{r, warning=F}
# RFE

# Define the RFE control settings
ctrl <- caret::rfeControl(functions = lrFuncs, # uses logistic regression as the underlying model
                   method = "cv",
                   number = 5)

# specify how many features RFE should try
sizes <- c(5, 10, 15, 20, 30)

set.seed(123)

# y needs to be a factor in `rfe()`
train$y <- as.factor(train$y)

rfe_model <- rfe(
    x = train[, -1],
    y = train$y,
    sizes = sizes,
    rfeControl = ctrl
)

rfe_model

rfe_features <- predictors(rfe_model)
rfe_features
```

```{r}
formula_rfe <- as.formula(
    paste("y ~", paste(rfe_features, collapse = "+"))
)

rfe_glm <- glm(formula_rfe, data = train, family = "binomial")

pred <- predict(rfe_glm, test, type = "response")
rfe_auc <- pROC::roc(test$y, pred)$auc
rfe_auc
```

## Embedded methods

### Lasso 

```{r}
# Lasso 
library(glmnet)

x_train <- model.matrix(y ~ ., train)[,-1]
y_train <- train$y

cv_lasso <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 1)

plot(cv_lasso)

lasso_coef <- coef(cv_lasso, s = "lambda.min")
lasso_features <- rownames(lasso_coef)[lasso_coef[,1] != 0]
lasso_features
```

```{r}
x_test <- model.matrix(y ~ ., test)[,-1]
pred <- predict(cv_lasso, s = "lambda.1se", newx = x_test, type = "response")

lasso_auc <- pROC::roc(test$y, pred)$auc
lasso_auc
```

### Gradient Boosting Trees

Gradient Boosting Trees is an ensemble learning method that builds a model by combining many small decision trees in a sequential way.

Instead of fitting a single large tree, the model builds trees one at a time, where each new tree focuses on correcting the mistakes made by the previous model.

#### Core idea

The model is built in an additive form:

$$\hat{f}(x) = \sum_{m=1}^{M} T_m(x)$$

where each $T_m(x)$ is a small decision tree. Each new tree is added to improve the overall prediction.

#### How it works step-by-step

#### 1. Start with a simple prediction
The model begins with an initial guess:

- For regression: the mean of $y$
- For classification: the log-odds of the outcome

#### 2. Compute errors (residuals)
The model calculates the difference between observed and predicted values:

$$\text{error} = y - \hat{y}$$

These errors represent what the model has not yet learned.

#### 3. Fit a tree to the errors
A small decision tree (weak learner) is trained to predict these residuals.

Importantly:

- The tree does not predict $y$ directly
- It predicts the remaining error

#### 4. Update the model
The new tree is added to improve predictions:

$$\hat{y}_{new} = \hat{y}_{old} + \eta \cdot T_m(x)$$

where: 

- $\eta$ is the learning rate (controls step size)
- $T_m(x)$ is the new tree

#### 5. Repeat the process
Steps 2–4 are repeated many times:

- Each tree corrects previous mistakes
- The model gradually improves

The XGBoost package implements a fast gradient-boosted decision tree algorithm. The model builds 100 sequential decision trees. 

Then we compute how important each feature is based on:

- how often it is used in splits
- how much it improves model performance.

```{r}
# Boosting 
library(xgboost)

xgb_model <- xgboost(x = x_train,
                     y = y_train,
                     nrounds = 100, # number of boosting iterations (trees)
                     objective = "binary:logistic")

importance_matrix <- xgb.importance(model = xgb_model)

xgb_features <- importance_matrix$Feature[1:10] 
xgb_features
```

```{r}
pred <- predict(xgb_model, x_test)

xgb_auc <- pROC::roc(test$y, pred)$auc
xgb_auc
```

# Summary

```{r}
# =========================
# SUMMARY OF RESULTS
# =========================

results <- data.frame(
  Method = c(
    "Baseline (all features)",
    "Filter (p-value)",
    "Filter (Mutual Information)",
    "Wrapper (Stepwise)",
    "Wrapper (RFE)",
    "Embedded (LASSO)",
    "Embedded (XGBoost)"
  ),
  
  AUC = c(
    baseline_auc,
    filter_auc,
    mi_auc,
    step_auc,
    rfe_auc,
    lasso_auc,
    xgb_auc
  ),
  
  Selected_Variables = c(
    "All features",
    paste(filter_features, collapse = ", "),
    paste(mi_features, collapse = ", "),
    paste(names(coef(step_model))[coef(step_model) != 0], collapse = ", "),
    paste(rfe_features, collapse = ", "),
    paste(lasso_features, collapse = ", "),
    paste0("Variable importance (no explicit subset) - top 10:", paste0(xgb_features, collapse = ", "))
  )
)
```

```{r}
results %>% arrange(desc(AUC))
```

# Discussion

- Which method achieved best predictive performance?

- Which features appear in every method?

